Efficient and Scalable Bipartite Matching with Fast Beta Linkage (fabl)

### 2021-09-27 methods_paper/fabl-paper.tex
update and notes for brian.

\section{Bipartite Record Linkage and Related Work}
As already mentioned, bipartite record linkage is the traditional set up studied in record linkage, where we assume there are two databases $\bm{X}_1$ and $\bm{X}_2$ such that there are duplications across the databases but not within each databases. We assume that each database has $n_1$ and $n_2$ records, and that $n_1 \leq n_2.$ Denote the number of entities simultaneously recorded by \emph{both databases} by $n_{12}$ such that $0 \leq n_{12} \leq n_2.$ Given this set up, as explained by \cite{steorts_bayesian_2016} and \cite{sadinle_bayesian_2017}, this can be viewed as bipartite record linkage. This is important as it allows one to view the matching matrix or linkage structure in the following way:
\begin{align}
\Delta_{ij} =
\begin{cases}
1 \quad \text{if records}\;  i \in \bm{X}_1 \; \text{and}\; j\in \bm{X}_2 \; \text{refer to the same entity}; \\
0 \quad \text{otherwise}.\\
\end{cases}
\end{align}
Bipartite matching can also be viewed as a matching labeling $\bm{Z} = (Z_1, \ldots, Z_{n_2})$ for the records in database $\bm{X}_2$ such that 
\begin{align}
Z_{j} =
\begin{cases}
i \quad \text{if records}\;  i \in \bm{X}_1 \; \text{and}\; j\in \bm{X}_2 \; \text{refer to the same entity}; \\
n_1 + j \quad n_1 + j \quad \text{if records}\;  j \in \bm{X}_2 \; \text{does not have a match in database}\; \bm{X}_1. \\
\end{cases}
\end{align}
Depending on which representation is the most convenient, we can go back and forth between the two using $\Delta_{ij} = I(Z_j = i),$ where $I(\cdot)$ is the indicator function.

@@

\begin{align}
P(\bm{Z}, \Phi \mid \bm{\Gamma}) = \prod_{j=1}^{n_1}  \prod_{i=1}^{n_2}\left[ \prod_{f=1}^{F}\prod_{l=1}^{L_f} m_{fl}^{I(Z_j = i)}u_{fl}^{I(Z_j \neq i)}\right]^{I(\gamma_{ij}^f = l)}, 
\end{align}
where $m_{f\ell} = P(\Gamma_{ij}^f = \ell \mid Z_j = i)$ denotes the probability of a match have level $\ell$ of disagreement in feature $f$ and 
$u_{f\ell} = P(\Gamma_{ij}^f = \ell \mid Z_j \neq i)$ represents the same probability for non-matches. 
In addition, denote $\bm{m}_f = (m_{f1}, \ldots, m_{fL_f})$, $\bm{u}_f = (u_{f1}, \ldots, u_{fL_f})$, 
$\bm{m}= (m_1, \ldots m_F)$, $\bm{u}= (u_1, \ldots u_F)$, and $\Phi = (\bm{m}, \bm{u})$.

@@

The likelihood becomes 
\begin{align}
P(\bm{Z}, \Phi \mid \bm{\gamma}^{obs}) = 
\prod_{j=1}^{n_1}  \prod_{i=1}^{n_2}\left[ \prod_{f=1}^{F}\prod_{l=1}^{L_f} m_{fl}^{a_{f\ell}(\bm{Z})}u_{fl}^{a_{f\ell}(\bm{Z})}\right], 
\end{align}
where 
\begin{align}
&a_{f\ell}(\bm{Z}) = \sum_{i,j} I_{obs}(\gamma_{ij}^f)I(\gamma_{ij}^f = \ell) I(Z_j = i) \\
&b_{f\ell}(\bm{Z}) = \sum_{i,j} I_{obs}(\gamma_{ij}^f) I(\gamma_{ij}^f = \ell) I(Z_j \neq i),
\end{align}
where $I_{obs}$ denotes if a value is observed or not. Next, for a particular matching configuration $\bm{Z}$, $a_{f\ell}(\bm{Z})$
and $b_{f\ell}(\bm{Z})$ denote the number of matches and non-matches with observed disagreement level $\ell$ for comparison feature $f.$ Using this parameterization, we specify independent conjugate priors for all features $f$:
\begin{align}
&\bm{m}_f \sim \text{Dirichlet}(\alpha_{f0}, \ldots, \alpha_{f L_f}) \\
&\bm{u}_f \sim \text{Dirichlet}(\beta_{f0}, \ldots, \beta_{f L_f}).
\end{align}

@@

In this section, we deviate from the approach of \cite{}, and propose a more scalable prior on the matching labelings $\bm{Z}.$ 
While the rest of the model has been proposed previously, to our knowledge, this particular prior has not been proposed before.

\textcolor{red}{Lay out Brian's prior here.}

@@

To indicate matching status, we adopt the \emph{linkage structure
	parameter} \(\mathbf{Z} = (Z_1, \ldots, Z_{n_B})\) from Sadinle 2017,
defined as \[Z_j=\begin{cases} 
	i,  & \text{if records } i\in A \text{ and } j\in B \text{ refer to the same entity}; \\
	n_A + 1,  & \text{if record } j\in B \text{ does not have a match in file } A; \\
\end{cases}\] This parameter is vector of length $n_B$, providing more memory efficient storage for the
linkage information than a \(n_A \times n_B\) sparse matrix of
indicators.

\textcolor{red}{Brian: This does not match what Sadinle has, so this needs to be justified regarding the change or if this is a typo. Could you please check this?}

\textcolor{blue}{Define \(m^{f\ell}:= P(\Gamma_{ij}^f = l |Z_j = i)\) as  the probability of a match having level $\ell$ of disagreement for feature $f.$ Define \(u^{f\ell}:= P(\gamma_{ij}^f = l |Z_j \neq i)\) as the corresponding probability for non-matches. Next, we adopt Fellegi and Sunter's conditionally independent fields assumption
that the level of agreement on one feature is independent of the level of
agreement on another.\footnote{Though this assumption is often not reasonable
(for example, two individuals of a different genders are more likely to have different names than two individuals of the same gender),
but it is common within the record linkage literature and generally
leads to models that perform well in practice; see discussion for
further remarks.} Denote \(\lambda\) to be the (marginal)
probability that some record \(j \in \bm{X}_2\) has a match in \(\bm{X}_1\). In addition, denote $\bm{m}_f = (m_{f1}, \ldots, m_{fL_f})$, $\bm{u}_f = (u_{f1}, \ldots, u_{fL_f})$, 
$\bm{m}= (m_1, \ldots m_F)$, $\bm{u}= (u_1, \ldots u_F)$, and $\Phi = (\bm{m}, \bm{u})$}.

\begin{enumerate}
\item Let's first write out the likelihood, which should be the same as in Sadinle
\item Then we should write out the Beta Prior, which should deviate and needs its own section in my opinion so the reader understands why it's a deviation
\item My understanding, is that section 4.1 and section 4.2 should look as they are, so these just need to be explained to the reader. The innovation is in the beta prior and making it more efficient. 
\end{enumerate}

@@

\hypertarget{model-specification}{%
	\subsection{Model Specification}
	\label{model-specification}}

For fields \(f \in \{1, \ldots, F\}\) and levels
\(l\in \{1, \ldots, L_f\}\) we adopt the following likelihood and prior
distributions.

$$P(\Gamma|\mathbf{Z}, \mathbf{m}, \mathbf{u}) = \prod_{j=1}^{n_B}  \prod_{i=1}^{n_A}\left[ \prod_{f=1}^{F}\prod_{l=1}^{L_f} m_{fl}^{I(Z_j = i)}u_{fl}^{I(Z_j \neq i)}\right]^{I(\gamma_{ij}^f = l)}$$

\textcolor{red}{It seems to me that this should be $P(\bm{Z}, \Phi \mid \bm{\Gamma})$ above and not what is currently written. }

\textcolor{red}{Are you handling missing values the way that Sadinle does? This isn't clear. It would he helpful to follow the Sadinle paper regarding structure as closely as possible given that it's very clear. I might recommend that you work on some details here regarding pages 12---13 as some important details are skipped over regarding what's random and observed that may not be obvious to the AE/reviewer.}

\[\mathbf{m^{f}} \sim \text{Dirichlet}(\alpha^{f1}, \ldots, \alpha^{fL_f})\]
\[\mathbf{u^{f}} \sim \text{Dirichlet}(\beta^{f1}, \ldots, \beta^{fL_f})\]
\[Z_j | \lambda =
\begin{cases} 
	\frac{1}{n_A}\lambda  & z_j \leq n_A; \\
	1-\lambda &  z_j  = n_A + 1 \\
\end{cases}\]

\[\lambda \sim \text{Beta}(\alpha_{\lambda}, \beta_{\lambda}) \] The
prior for \(Z_j\) has equal probability of matching to all records
\(i\in A\), and non-matching probability governed by \(\lambda\).
Therefore a \(\lambda \sim \text{Beta}(1, 1)\) corresponds to a prior
belief that nonmatches and matches are equally likely, and a
\(\lambda \sim \text{Beta}\left(1, \frac{1}{n_A}\right)\) prior
corresponds to a uniform prior on the labeling of \(\mathbf{Z}\).

>>>>

\section{Bipartite Record Linkage and Related Work}
As already mentioned, bipartite record linkage is the traditional set up studied in record linkage, where we assume there are two databases $\bm{X}_1$ and $\bm{X}_2$ such that there are duplications across the databases but not within each databases. We assume that each database has $n_1$ and $n_2$ records, and that $n_1 \leq n_2.$ Denote the number of entities simultaneously recorded by \emph{both databases} by $n_{12}$ such that $0 \leq n_{12} \leq n_2.$ Given this set up, as explained by \cite{steorts_bayesian_2016} and \cite{sadinle_bayesian_2017}, this can be viewed as bipartite record linkage. This is important as it allows one to view the matching matrix or linkage structure in the following way:
\begin{align}
\Delta_{ij} =
\begin{cases}
1 \quad \text{if records}\;  i \in \bm{X}_1 \; \text{and}\; j\in \bm{X}_2 \; \text{refer to the same entity}; \\
0 \quad \text{otherwise}.\\
\end{cases}
\end{align}
Bipartite matching can also be viewed as a matching labeling $\bm{Z} = (Z_1, \ldots, Z_{n_2})$ for the records in database $\bm{X}_2$ such that 
\begin{align}
Z_{j} =
\begin{cases}
i \quad \text{if records}\;  i \in \bm{X}_1 \; \text{and}\; j\in \bm{X}_2 \; \text{refer to the same entity}; \\
n_1 + j \quad  \quad \text{if records}\;  j \in \bm{X}_2 \; \text{does not have a match in database}\; \bm{X}_1. \\
\end{cases}
\end{align}
Depending on which representation is the most convenient, we can go back and forth between the two using $\Delta_{ij} = I(Z_j = i),$ where $I(\cdot)$ is the indicator function.

@@

\begin{align}
\mathcal{L}(\bm{Z}, \Phi \mid \bm{\Gamma}) = \prod_{j=1}^{n_1}  \prod_{i=1}^{n_2}\left[ \prod_{f=1}^{F}\prod_{l=1}^{L_f} m_{fl}^{I(Z_j = i)}u_{fl}^{I(Z_j \neq i)}\right]^{I(\gamma_{ij}^f = l)}, 
\end{align}
where $m_{f\ell} = P(\Gamma_{ij}^f = \ell \mid Z_j = i)$ denotes the probability of a match have level $\ell$ of disagreement in feature $f$ and 
$u_{f\ell} = P(\Gamma_{ij}^f = \ell \mid Z_j \neq i)$ represents the same probability for non-matches. 
In addition, denote $\bm{m}_f = (m_{f1}, \ldots, m_{fL_f})$, $\bm{u}_f = (u_{f1}, \ldots, u_{fL_f})$, 
$\bm{m}= (m_1, \ldots m_F)$, $\bm{u}= (u_1, \ldots u_F)$, and $\Phi = (\bm{m}, \bm{u})$.

@@

The likelihood becomes 
\begin{align}
\mathcal{L}(\bm{Z}, \Phi \mid \bm{\gamma}^{obs}) = 
\prod_{j=1}^{n_1}  \prod_{i=1}^{n_2}\left[ \prod_{f=1}^{F}\prod_{l=1}^{L_f} m_{fl}^{a_{f\ell}(\bm{Z})}u_{fl}^{a_{f\ell}(\bm{Z})}\right], 
\end{align}
where 
\begin{align}
&a_{f\ell}(\bm{Z}) = \sum_{i,j} I_{obs}(\gamma_{ij}^f)I(\gamma_{ij}^f = \ell) I(Z_j = i) \\
&b_{f\ell}(\bm{Z}) = \sum_{i,j} I_{obs}(\gamma_{ij}^f) I(\gamma_{ij}^f = \ell) I(Z_j \neq i),
\end{align}
where $I_{obs}$ denotes if a value is observed or not. Next, for a particular matching configuration $\bm{Z}$, $a_{f\ell}(\bm{Z})$
and $b_{f\ell}(\bm{Z})$ denote the number of matches and non-matches with observed disagreement level $\ell$ for comparison feature $f.$ Using this parameterization, we specify independent conjugate priors for all features $f$:
\begin{align}
&\bm{m}_f \sim \text{Dirichlet}(\alpha_{f0}, \ldots, \alpha_{f L_f}) \\
&\bm{u}_f \sim \text{Dirichlet}(\beta_{f0}, \ldots, \beta_{f L_f}).
\end{align}

@@

In this section, we deviate from the approach of \cite{sadinle_bayesian_2017}, and propose a more scalable prior on the matching labelings $\bm{Z} = (Z_1, Z_2, \ldots Z_{n_2}),$ where recall $Z_j \in \{1,2, 3, n_1, n_1 + j\}.$

While the rest of the model has been proposed previously, to our knowledge, this particular prior has not been proposed before.

\textcolor{red}{Brian: could you please lay out your proposal here more clear regarding the notation above and add as much details as possible regarding your approach. More detail is better than less. It will be important to contrast this with the prior that Sadinle uses and note that you're dropping the constraint. (Make sure to include any proofs/derivations that might not be obvious to a reader given this is a major contribution to the paper).}

\[Z_j | \lambda =
\begin{cases} 
	\frac{1}{n_1}\lambda  & z_j \leq n_1; \\
	1-\lambda &  z_j  = n_1 + j \\
\end{cases}\]

\[\lambda \sim \text{Beta}(\alpha_{\lambda}, \beta_{\lambda}) \] The
prior for \(Z_j\) has equal probability of matching to all records
\(i\in A\), and non-matching probability governed by \(\lambda\).
Therefore a \(\lambda \sim \text{Beta}(1, 1)\) corresponds to a prior
belief that nonmatches and matches are equally likely, and a
\(\lambda \sim \text{Beta}\left(1, \frac{1}{n_1}\right)\) prior
corresponds to a uniform prior on the labeling of \(\mathbf{Z}\).

@@

%To indicate matching status, we adopt the \emph{linkage structure
%	parameter} \(\mathbf{Z} = (Z_1, \ldots, Z_{n_B})\) from Sadinle 2017,
%defined as \[Z_j=\begin{cases} 
%	i,  & \text{if records } i\in A \text{ and } j\in B \text{ refer to the same entity}; \\
%	n_A + 1,  & \text{if record } j\in B \text{ does not have a match in file } A; \\
%\end{cases}\] This parameter is vector of length $n_B$, providing more memory efficient storage for the
%linkage information than a \(n_A \times n_B\) sparse matrix of
%indicators.
%
%\textcolor{red}{Brian: This does not match what Sadinle has, so this needs to be justified regarding the change or if this is a typo. Could you please check this?} 
%
%
%\textcolor{blue}{Define \(m^{f\ell}:= P(\Gamma_{ij}^f = l |Z_j = i)\) as  the probability of a match having level $\ell$ of disagreement for feature $f.$ Define \(u^{f\ell}:= P(\gamma_{ij}^f = l |Z_j \neq i)\) as the corresponding probability for non-matches. Next, we adopt Fellegi and Sunter's conditionally independent fields assumption
%that the level of agreement on one feature is independent of the level of
%agreement on another.\footnote{Though this assumption is often not reasonable
%(for example, two individuals of a different genders are more likely to have different names than two individuals of the same gender),
%but it is common within the record linkage literature and generally
%leads to models that perform well in practice; see discussion for
%further remarks.} Denote \(\lambda\) to be the (marginal)
%probability that some record \(j \in \bm{X}_2\) has a match in \(\bm{X}_1\). In addition, denote $\bm{m}_f = (m_{f1}, \ldots, m_{fL_f})$, $\bm{u}_f = (u_{f1}, \ldots, u_{fL_f})$, 
%$\bm{m}= (m_1, \ldots m_F)$, $\bm{u}= (u_1, \ldots u_F)$, and $\Phi = (\bm{m}, \bm{u})$}.
%
%\begin{enumerate}
%\item Let's first write out the likelihood, which should be the same as in Sadinle
%\item Then we should write out the Beta Prior, which should deviate and needs its own section in my opinion so the reader understands why it's a deviation
%\item My understanding, is that section 4.1 and section 4.2 should look as they are, so these just need to be explained to the reader. The innovation is in the beta prior and making it more efficient. 
%\end{enumerate}

@@

%\hypertarget{model-specification}{%
%	\subsection{Model Specification}
%	\label{model-specification}}
%
%For fields \(f \in \{1, \ldots, F\}\) and levels
%\(l\in \{1, \ldots, L_f\}\) we adopt the following likelihood and prior
%distributions.

%$$P(\Gamma|\mathbf{Z}, \mathbf{m}, \mathbf{u}) = \prod_{j=1}^{n_B}  \prod_{i=1}^{n_A}\left[ \prod_{f=1}^{F}\prod_{l=1}^{L_f} m_{fl}^{I(Z_j = i)}u_{fl}^{I(Z_j \neq i)}\right]^{I(\gamma_{ij}^f = l)}$$

%\textcolor{red}{It seems to me that this should be $P(\bm{Z}, \Phi \mid \bm{\Gamma})$ above and not what is currently written. }
%
%\textcolor{red}{Are you handling missing values the way that Sadinle does? This isn't clear. It would he helpful to follow the Sadinle paper regarding structure as closely as possible given that it's very clear. I might recommend that you work on some details here regarding pages 12---13 as some important details are skipped over regarding what's random and observed that may not be obvious to the AE/reviewer.}
%
%\[\mathbf{m^{f}} \sim \text{Dirichlet}(\alpha^{f1}, \ldots, \alpha^{fL_f})\]
%\[\mathbf{u^{f}} \sim \text{Dirichlet}(\beta^{f1}, \ldots, \beta^{fL_f})\]
